A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-03-16 CC/writeup.Rnw
write-up as of 6:45 on Monday.  new data was also added to dropbox.

\documentclass[11pt]{article}
\usepackage{asa}
\usepackage{xspace}
\usepackage{natbib}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
% \geometry{landscape} % rotated page geometry
% See the ``Article customise'' template for come common customisations

@@

@@

\par The \texttt{GGally} package by \citet{ggally} contains a very useful function written by Fran\c{c}ois Briatte and Moritz Marbach called \texttt{ggnet} that does allow for fairly straightforward customization of these three necessary graph attributes. \hh{XXX drive home the message that vertices, edges and layout are important, and repeat at this point} \sctq{Not quite sure what to say here. What follows is my guesstimate.} \sct{Coloring the vertices or edges in a graph is a quick and easy way to visualize grouping and can help with pattern or cluster detection. The vertices in a network and the edges between them compose the structure of a network, and being able to discover patterns among them visually is a key part of network analysis. Viewing multiple layouts of the same network can also help reveal patterns or clusters that would not be discovered when only viewing one layout or analyzing only an adjacency matrix.} The \texttt{ggnet} function, however, requires the graph input to be a \texttt{network} object according to the \texttt{network} package.  Our goal is to build off of the \texttt{ggnet} function and present an intuitive \texttt{geom} for network visualization within the \texttt{ggplot2} framework, without the need for objects other than simple data frames.  
%
  \subsection{Data Structure}
  In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. \emph{The only necessary variables are the vertex names/labels and the vertex degree (if undirected) or the vertex in degree and out degree (if directed).  The names and degrees are stored in columns and the rows are the vertices.  Other values of interest, such as grouping variables, for each vertex should be stored as columns in this dataframe, with an observation for each vertex.}  
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should be listed according to the names of the vertices in the vertex information data frame. Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.%\footnote{I'm not 100\% certain of all of this yet, but so far it seems to me to be the best way to think about networks in terms of data frames.}

@@

\section{Many Many Examples}
\section{Technical Details}
\sctq{I have no idea how detailed I should be. I'm going to attempt to be detailed without parsing each line in the code, but I do want to describe how everything works in detailed enough descriptions. I know you'll tell me if I'm too detailed or not detailed enough!}
\sct{
In order to construct the \texttt{geom} in \texttt{ggplot2}, we needed two \texttt{R} script files: a `stat' file and a `geom' file. The stat file performs all the necessary calculations on the two input data frames, and the geom file performs the drawing of the network.
  \subsection{Creating \texttt{stat\_net}}
  In order to plot the network with only the edge connections and vertex names, we relied on the \texttt{sna} and \texttt{network} packages \citep{sna, network.jss}. First, we used the edges data frame to construct a network object with the \texttt{as.network()} function, then constructed an adjacency matrix from that network using the \texttt{as.matrix.network.adjacency()} function.  This adjacency matrix is then passed to the chosen \texttt{gplot.layout.*()} function from the \texttt{sna} package.  \emph{When the layout argument is changed in the \texttt{geom\_net()} function, the corresponding \texttt{gplot.layout.*()} function in \texttt{sna} is called with a \texttt{do.call()} statement.} This layout function produces a matrix of coordinates of the vertices, which we then transform into a data frame containing the coordinates of the edges using the \texttt{as.matrix.network.edgelist()} function in \texttt{network}.  We are grateful to Moritz Marbach for his \texttt{gplot()} function which we used when creating this process \citep{marbach}.  At the end of the previous routine, we have a data frame with \texttt{x, y, xend,} and \texttt{yend} columns describing the edges and another data frame with \texttt{x,y} columns describing the vertices.  These two data frames, along with the many \texttt{aes()} and other argument options get passed on to the actual network geometry for drawing. 
  \subsection{Creating \texttt{geom\_net}}
  We constructed \texttt{geom\_net} using the drawing capabilities of the \texttt{geom\_point}, \texttt{geom\_line}, \emph{and \texttt{geom\_text}} functions that were already in \texttt{ggplot2}.  Using these capabilities, we were able to seamlessly incorporate our network geometry into the \texttt{ggplot2} structure.   We took our edges data frame and used that as our input to the \texttt{GeomSegment} draw function.  Then, on top of that, we plot the vertices by passing our vertex data frame to the \texttt{GeomPoint} draw function. \emph{Finally, we use the \texttt{GeomText} draw function to add the vertex and edge labels when the \texttt{vlabel} and \texttt{elabel} arguments, respectively, are true. *insert more detail here after figuring it out*}
} \sctq{This doesn't seem very long, but I already feel like there's too much detail! Should I add more?}

>>>>

\documentclass[11pt]{article}
\usepackage{asa}
\usepackage{xspace}
\usepackage{natbib}
\usepackage{url} 
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
% \geometry{landscape} % rotated page geometry
% See the ``Article customise'' template for come common customisations

@@

% <<readingdata, message=FALSE>>=
%  library(ggplot2)
%  theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
%  data(blood, football, theme_elements)
%  email.net <- read.csv("~/Dropbox/ggplot2/email-data.csv") 
%  email.net <- subset(email.net, day %in% c(9,13))
%  email.names <- data.frame(name = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to))))) 
% @

@@

\par The \texttt{GGally} package by \citet{ggally} contains a very useful function written by Fran\c{c}ois Briatte and Moritz Marbach called \texttt{ggnet} that does allow for fairly straightforward customization of these three necessary graph attributes. \hh{XXX drive home the message that vertices, edges and layout are important, and repeat at this point} \sctq{Not quite sure what to say here. What follows is my guesstimate.} \sct{Coloring the vertices or edges in a graph is a quick and easy way to visualize grouping and can help with pattern or cluster detection. The vertices in a network and the edges between them compose the structure of a network, and being able to discover patterns among them visually is a key part of network analysis. Viewing multiple layouts of the same network can also help reveal patterns or clusters that would not be discovered when only viewing one layout or analyzing only an adjacency matrix.} The \texttt{ggnet} function, however, requires the graph input to be a \texttt{network} object according to the \texttt{network} package.  Our goal is to build off of the \texttt{ggnet} function and present an intuitive \texttt{geom} for network visualization within the \texttt{ggplot2} framework, without the need for objects other than simple data frames.

  \subsection{Data Structure}
  In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. \emph{The only necessary variables are the vertex names/labels and the vertex degree (if undirected) or the vertex in degree and out degree (if directed).  The names and degrees are stored in columns and the rows are the vertices.  Other values of interest, such as grouping variables, for each vertex should be stored as columns in this dataframe, with an observation for each vertex.}  
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should be listed according to the names of the vertices in the vertex information data frame. Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.%\footnote{I'm not 100\% certain of all of this yet, but so far it seems to me to be the best way to think about networks in terms of data frames.}

@@

\section{Many Many Examples}
\sctq{Format-wise, is it ok if each example is its own subsection?}\\
\sct{
In this section, we demonstrate the capabilities of \texttt{geom\_net} in a series of examples. 
  \subsection{Blood Donation}
  \sctq{I had to comment out all the code chunks for now. Not sure why but it won't compile if they're there.}
  \emph{Note: would be best with labels and direction}
% # <<blood_ex>>=
% # ggplot(data = blood$edges, aes(from_id = from, to_id = to)) + geom_net(vertices = blood$vertices) + expand_limits(x = c(-5,5), y = c(-5,5)) + theme_net #add vlabel later
% # @
  \subsection{Email Network}
  \emph{Note: would be best with labels and direction}
% # <<email_ex>>=
% # ggplot(data = email.net, aes(from_id = From, to_id = to, ealpha = 0.1,vsize = 2)) + geom_net(vertices = email.names) + expand_limits(x=c(-1,5), y = c(-1,5)) + facet_wrap(~day) + theme_net
% # @
  \subsection{\texttt{ggplot2} Theme Elements}
  \emph{Note: would be best with labels and direction}
% # <<theme_ex>>=
% # #ggplot(data = theme_elements$edges, aes(from_id = child, to_id = parent)) + geom_net(vertices = theme_elements$vertices) + expand_limits(x = c(-5,5), y = c(-5,5)) + theme_net
% # @
  \subsection{Madmen Network (again?? or no?)}
  \sctq{leaving blank for now.}
  \subsection{College Football}
  \emph{Note: would be best with coloring vertices by conference} \citet{football}
% # <<football_ex>>=
% # ggplot(data = football$edges, aes(from_id = from, to_id = to)) + geom_net(vertices = football$vertices) + expand_limits(x = c(-2.5,2.5), y = c(-2.5,2.5)) + theme_net
% # @
  \subsection{Les Mis\'{e}rables}
  \emph{Haven't looked at this one yet. Just found it on MEJ Newmans website.  http://www-personal.umich.edu/~mejn/netdata/}
}

\section{Technical Details}
\sctq{I have no idea exactly how detailed I should be. I'm going to attempt to be detailed without parsing each line in the code, but I do want to describe how everything works in detailed enough descriptions. I know you'll tell me if I'm too detailed or not detailed enough!}
\sct{
In order to construct the \texttt{geom} in \texttt{ggplot2}, we needed two \texttt{R} script files: a `stat' file and a `geom' file. The stat file performs all the necessary calculations on the two input data frames, and the geom file performs the drawing of the network.
  \subsection{Creating \texttt{stat\_net}}
  In order to plot the network with only the edge connections and vertex names, we relied on the \texttt{sna} and \texttt{network} packages \citep{sna, network.jss}. First, we used the edges data frame to construct a network object with the \texttt{as.network()} function, then constructed an adjacency matrix from that network using the \texttt{as.matrix.network.adjacency()} function.  This adjacency matrix is then passed to the chosen \texttt{gplot.layout.*()} function from the \texttt{sna} package.  \emph{When the layout argument is changed in the \texttt{geom\_net()} function, the corresponding \texttt{gplot.layout.*()} function in \texttt{sna} is called with a \texttt{do.call()} statement.} This layout function produces a matrix of coordinates of the vertices, which we then transform into a data frame containing the coordinates of the edges using the \texttt{as.matrix.network.edgelist()} function in \texttt{network}.  We are grateful to Moritz Marbach for his \texttt{gplot()} function which we used when creating this process \citep{marbach}.  At the end of the previous routine, we have a data frame with \texttt{x, y, xend,} and \texttt{yend} columns describing the edges and another data frame with \texttt{x,y} columns describing the vertices.  These two data frames, along with the many \texttt{aes()} and other argument options get passed on to the actual network geometry for drawing. 
  \subsection{Creating \texttt{geom\_net}}
  We constructed \texttt{geom\_net} using the drawing capabilities of the \texttt{geom\_point}, \texttt{geom\_line}, \emph{and \texttt{geom\_text}} functions that were already in \texttt{ggplot2}.  Using these capabilities, we were able to seamlessly incorporate our network geometry into the \texttt{ggplot2} structure.   We took our edges data frame and used that as our input to the \texttt{GeomSegment} draw function.  Then, on top of that, we plot the vertices by passing our vertex data frame to the \texttt{GeomPoint} draw function. \emph{Finally, we use the \texttt{GeomText} draw function to add the vertex and edge labels when the \texttt{vlabel} and \texttt{elabel} arguments, respectively, are true. *insert more detail here after figuring it out*}
} \sctq{This doesn't seem very long, but I already feel like there's too much detail! Should I add more?}
